% A workaround to allow relative paths in included subfiles
% that are to be compiled separately
% See https://tex.stackexchange.com/questions/153312/subfiles-inside-a-subfile-using-relative-paths
\providecommand{\main}{..}
\documentclass[\main/thesis.tex]{subfiles}

\begin{document}
\begin{acknowledgements}
  First and foremost, I would like to thank my supervisor, J. Nelson Amaral, whose constant support has helped me finish the largest undertaking yet seen in my life.
  His passion for teaching has helped foster my love for all things computers more than I could have ever known.

  Collectively, I would like to thank Jo\~ao Paulo Labegalini de Carvalho and Ivan Korostolev for being companions throughout my time as a student.
  They were always ready to be a sounding board for new ideas and offered encouragement and assistance when solving difficult problems.
  I looked forward to my time in the lab thanks to these two.

  I would also like to thank the team at IBM who supported my research, notably Kit Barton and Jose Moreira, both of whom were essential in the completion of this thesis.
  Every resource and bit of knowledge they had was made available with a simple message.
  I must also acknowledge the backend team, led by Nemanja Ivanovic, for their aid and their tolerance as I played with and broke the things under their care.

  Funding was necessary to complete the work contained in this thesis; without support from the IBM Canada Centre for Advanced Studies and the Natural Sciences and Engineering Research Council of Canada through their Collaborative Research and Development program, none of this would have been possible.
\end{acknowledgements}
\end{document}
